Inferring Invariants of Distributed Programs

### 2015-04-30 Doc/approach.tex
Background revisited

The ultimate goal of this project is to infer invariants in a
distributed system. For the scope of this project we will consider a
system with 2 nodes. Before inferring invariants we need to identify
likely distributed state variables. We define distributed state
variables, as variables that are stored in one node whose values are
affected or determined by the other remote node in the distributed
system. It should be noted that these variables are only a part of the
state in a distributed system at a given time. For our case we assume
that these capture a significant portion of the state. Consider the
code samples Listing~\ref{lst:node0} and Listing~\ref{lst:node1}
showing the communication between two nodes in a distributed system.
This conversation is depicted visually in
Figure~\ref{fig:sample_code_diag}. Here it will be useful to to infer
the equality between elements in each pair \texttt{(a,first),
(b,second),(sum,result)} and also the properites: \texttt{(result = a
+ b)} and \texttt{(sum = first + second)}.

@@

The naive approach for inferring these invariants will be to dump all
state varaible values at all program points and feed them to an
invariant detector like daikon\cite{ernst2007daikon}. But given the
number of program points and huge number of variables in a relatively
large program this approach is not scalable and also we can end up inferring incorrect invariants that are hold by accident. So we propose two main
optimizations to reduce the number of program point, variable value
combinations.

@@

Rather than dumping state at all program points we expect the
developer to mark interesting program points using annotations. The
program state will be dumped at only these annotated points. This will
also help to narrow down the focus to points that are identified by
the developer as interesting points without wasting resources
analyzing useless program points. The annotations will be identified
by the parser at the instrumentation phase and re-written to include
statements which will write the values of variable to a file when the
program reaches that point in the runtime.

Since there can be multiple points annotated in a program by a
developer, vector clocks will be used to partially order annotation
pairs in the two programs.% This will also help to invalidate
% annotations inserted at irrelevant locations by the developer.

@@

Even at interesting program points, variable space that need to be
captured is very large and also we want to filter out irrelevant invariants by only considering variables that affect each other for invariant detection phase. So only the variables that were affected by
the communication with other nodes need to be identified. To detect
these variables, some form of dataflow analysis is required. This
includes the dataflow of the distributed system as a whole and also in
a single program. For example using a forward flow analysis on Listing~\ref{lst:node1} we can say that \texttt{sum} variable at line 4 and line 5 is affected by the \texttt{buf} variable at line 1. But currently there are no data or control flow analysis
frameworks for Go programs.

@@

``A program slice consists of the parts of a program that (potentially) affect the values computed at some point of interest"\cite{Tip}. For example figure \ref{Fig:backslice} illustrates an example Go program and its computed backward slice from line 11, the sliced program only contains statements that affect value of variables used at line 11 (i.e. product). To identify variables affected by the communication with other nodes we computed the forward slice from receive message functions and computed backward slice from send message functions. To compute these forward and backward slices we used the approach first proposed by Ottenstein and Ottenstein \cite{PDG}, which used Program Dependence Graph (PDG) and computed slices by checking the reachability of nodes in this graph. PDG captures both control dependency and data dependency of the program, in fact PDG is computed by combining both Data Dependence Graph and Control Dependence Graph (CDG) of a program. Figure \ref{fig:pdg} illustrates our approach to compute PDG. First, we computed the Control Flow Graph (CFG) of the program, then from the reversed CFG we computed the post-dominance tree. The Dominance tree is computed by the algorithm proposed by Lengauer and Tarjan in \cite{Dom}. In the next step we computed the CDG from post-dominance tree. To compute Data Dependence Graph, first we computed the reaching definitions by solving dataflow equations\cite{dragonbook} and then used $(Def,Use)$ pair to compute the data dependency between associated statements and form Data Dependence Graph. We combined Data and Control Dependence Graphs to compute the PDG. Having the PDG of the program we checked the reachability of nodes in PDG starting from start node to compute forward and backward slices.

@@

>>>>

The ultimate goal of this project is to infer invariants in a
distributed system. For this project we consider a
system with 2 nodes. Before inferring invariants we need to identify
likely distributed state variables. We define distributed state
variables, as variables that are stored in one node whose values are
affected or determined by the other remote node in the distributed
system. It should be noted that these variables are only a part of the
state in a distributed system at a given time. For our case we assume
that these capture a significant portion of the state. Consider the
code samples Listing~\ref{lst:node0} and Listing~\ref{lst:node1}
showing the communication between two nodes in a distributed system.
This conversation is depicted visually in
Figure~\ref{fig:sample_code_diag}. Here it will be useful to to infer
the equality between elements in each pair \texttt{(a,first),
(b,second),(sum,result)} and also the properties: \texttt{(result = a
+ b)} and \texttt{(sum = first + second)}.

@@

As mentioned earlier, the naive approach for inferring these invariants will be to dump all
state variable values at all program points and feed them to an
invariant detector like Daikon\cite{ernst2007daikon}. But given the
number of program points and huge number of variables in a relatively
large program this approach is not scalable. Moreover, it can result in inferring incorrect invariants that are hold by accident. So we propose two main
optimizations to reduce the number of program point, variable value
combinations.

@@

Rather than dumping state at all program points we expect the
developer to mark interesting program points using annotations. The
program state will be dumped at only these annotated points. This will
also help to narrow down the focus to points that are identified by
the developer as interesting points without wasting resources
analyzing useless program points. The annotations will be identified
by the parser at the instrumentation phase and re-written to include
statements which will write the values of variable to a file when the
program reaches that point in the runtime. Since there can be multiple points annotated in a program by a
developer, vector clocks will be used to partially order annotation
pairs in the two programs.% This will also help to invalidate
% annotations inserted at irrelevant locations by the developer.

@@

Even at interesting program points, variable space that need to be
captured is very large and also we want to filter out irrelevant invariants by only considering variables that affect each other for invariant detection phase. So only the variables that were affected by
the communication with other nodes need to be identified (distributed state variables).  To detect
these variables, some form of data and control flow analysis is required. For example using a backward flow analysis on Listing~\ref{lst:node1} we can say that \texttt{sum} variable at line 4 and line 5 is affected by the \texttt{buf} variable and receive function at line 1. Note that \texttt{sum} variable is part of distributed state of the system, since its value is affected by a receive function and affects a send function, so it is possibly affected and affects variables in other nodes of the system (here it is affected variables \texttt{a} and \texttt{b} and affects \texttt{result} in Node 0). To compute these distributed state variables, we developed a Go program slicer which we explain in the following section.

@@

``A program slice consists of the parts of a program that (potentially) affect the values computed at some point of interest"\cite{Tip}. For example figure \ref{Fig:backslice} illustrates an example Go program and its computed backward slice from line 11, the sliced program only contains statements that affect value of variables used at line 11 (i.e. product). To identify variables affected by the communication with other nodes we computed the forward slice from receive message functions and computed backward slice from send message functions and determined the variables assigned in these sliced statements. To compute these forward and backward slices we used the approach first proposed by Ottenstein and Ottenstein \cite{PDG}, which used Program Dependence Graph (PDG) and computed slices by checking the reachability of nodes in this graph. PDG captures both control dependency and data dependency of the program, in fact PDG is computed by combining both Data Dependence Graph and Control Dependence Graph (CDG) of a program. For example figure \ref{fig:expdg} depicts the PDG of example program of figure \ref{Fig:backslice}. Figure \ref{fig:pdg} illustrates our approach to compute PDG. First, we computed the Control Flow Graph (CFG) of the program, then from the reversed CFG we computed the post-dominance tree. The Dominance tree is computed by the algorithm proposed by Lengauer and Tarjan \cite{Dom}. In the next step we computed the CDG from post-dominance tree. To compute Data Dependence Graph, first we computed the reaching definitions by solving dataflow equations\cite{dragonbook} and then used $(Def,Use)$ pair to compute the data dependency between associated statements and form Data Dependence Graph. We combined Data and Control Dependence Graphs to compute the PDG. Having the PDG of the program we checked the reachability of nodes in PDG starting from start node to compute forward and backward slices. For example, In PDG of figure \ref{fig:expdg} thick lines are control dependence edges and thin lines are data dependence edges, starting from \texttt{send(product)} node and going backward shaded nodes are reachable, so the shaded nodes are in backward slice of \texttt{send(product)} statement.

@@

\begin{figure}
  \includegraphics[width=\columnwidth,scale=0.5]{images/pdg.png}
  \caption{Program Dependence Graph of Example Program in Figure \ref{Fig:backslice}}
  \label{fig:expdg}
\end{figure}
